---
title: "Hierarchical Models on Spotify Data"
subtitle: Hierarchical Models and Model Diagnostics
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Problem Overview

We are using the spotify dataset <https://rdrr.io/cran/bayesrules/man/spotify.html>. This dataset include 350 songs and 23 variables. In this problem, we try to use the song artist and the danceability of the song to predict the popularity. In below, I would explore using normal regression and hierarchical Bayesian modelling.

# Setup

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(infer)
library(broom)
library(rstan)
library(rstantools)
library(cowplot)
library(akima)
library(infer)
library(bayesplot)
library(bayesrules)
rstan_options(auto_write = TRUE)  # To save some compiling
```

\newpage

### 1. Data Wrangling

1.  We only need the column artist, danceability and popularity

```{r}
spotify_training <-spotify %>%
  select(artist, danceability, popularity)
spotify_training

```

```{r}
artist_catalogue <- tibble(
  artist = levels(spotify_training$artist),
  code = 1:44
)
```

### 2. Exploratory Data Analysis

```{r fig.width=11, fig.height=7, tags=c()}

scatter_plot <- ggplot(spotify_training, aes(x = danceability, y = popularity)) +
  geom_point() +  # Scatterplot
  labs(x = "Danceability", y = "Popularity") +  # Axes labels
  ggtitle("Scatterplot of Popularity vs Danceability") +  # Plot title
  theme_minimal()  # Minimal theme (adjust as needed)

scatter_plot 
```

```{r fig.width=20, fig.height=15, warning=FALSE, tags=c()}
popularity_boxplots <- spotify_training %>%
  ggplot(aes(reorder(artist, popularity), popularity)) +
  geom_boxplot(aes(fill = artist)) +
  scale_fill_viridis_d(option="plasma")+
  labs(y = "Popularity Score", x = "Artist") +
  ggtitle("Side-by-Side Boxplots of Popularity by Artist") +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 14, angle = 90),
    axis.title = element_text(size = 14),
    legend.position = "none"
  )

popularity_boxplots

```

### 3. Modelling with a linear regression

```{r}
regression_model <- lm(popularity ~ danceability + artist, data = spotify_training)
```

```{r}
summary(regression_model)
```

```{r}
# Scatterplot with regression lines by artist and smaller legend
scatter_plot <- ggplot(spotify_training, aes(x = danceability, y = popularity, color = artist)) +
  geom_point() +  # Scatterplot
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +  # Regression lines without confidence intervals
  labs(x = "Danceability", y = "Popularity", title = "Scatterplot of Popularity vs Danceability by Artist") +
  theme_minimal() +
  scale_color_discrete(name = "Artist") +  # Legend title
  theme(
    legend.text = element_text(size = 6)  # Adjust legend text size
  )

# Print the scatterplot with smaller legend
print(scatter_plot)


```

Here we see that for some artists there are less data points, and therefore our prediction is not doing very great, with a r-square pretty low. Also the prediction interval would be very huge because of our uncertainty.

### 4. Using Bayesian Model to model the data

$$
\begin{align*} \text{likelihood:} \qquad Y_{i,j} \mid \mu_j, \sigma_w, \beta, X_{i,j} \sim \mathcal{N}(\mu_j + \beta X_{i,j}, \sigma^2_w) \\ \text{where} \quad \mu_j = \mu + b_j \\ \text{priors:} \quad b_j \mid \sigma_\mu \sim \mathcal{N}(0, \sigma_\mu^2) \\ \qquad \mu \sim \mathcal{N}(50, 52^2) \\ \qquad \beta \sim \mathcal{N}(0, 1) \\ \qquad \sigma_w \sim \text{Exponential}(1) \\ \qquad \sigma_\mu \sim \text{Exponential}(0.048). \end{align*}
$$

In this model, $Y_{i,j}$ represents the popularity score of the $i$-th song by the $j$-th artist,

$X_{i,j}$ represents the danceability score of the $i$-th song by the $j$-th artist.

The popularity score of the **`i`**-th song is normally distributed with a mean that is the sum of the overall mean popularity, the artist-specific deviation in popularity, and the product of the danceability score and a coefficient. The standard deviation of this distribution is the within-artist standard deviation for popularity (**`within_sigma_popularity`**).

The mean $\mu_j$ for each artist is modeled as a global mean $\mu$ plus an artist-specific deviation $b_j$. The artist-specific deviations are assumed to be normally distributed with mean 0 and between-artist standard deviation $\sigma_\mu$.

The priors for the global mean $\mu$, the coefficient $\beta$ of the danceability score, the within-artist standard deviation $\sigma_w$, and the between-artist standard deviation $\sigma_\mu$ are also specified. In the prior, it is estimated that the between artist variability would be higher than the within artist variability.

### 4.1 Visualizing Priors

```{r}
prior_normal_50_52 <- ggplot() +
  xlim(-200, 300) +
  ylim(0, 0.01) +
  geom_function(fun = dnorm, args = list(mean = 50, sd = 52), linewidth = 1) +
  theme(
    plot.title = element_text(size = 16),
    axis.text.x = element_text(size = 12, angle = 0),
    axis.text.y = element_text(size = 12, angle = 0),
    axis.title = element_text(size = 12),
  ) +
  labs(y = "Density", x = expression(mu)) +
  ggtitle(expression(paste("Prior Normal(50, ", 52^2, ")")))

prior_exp_0.048 <- ggplot() +
  xlim(0, 140) +
  ylim(0, 0.05) +
  geom_function(fun = dexp, args = list(rate = 0.048), linewidth = 1) +
  theme(
    plot.title = element_text(size = 16),
    axis.text.x = element_text(size = 12, angle = 0),
    axis.text.y = element_text(size = 12, angle = 0),
    axis.title = element_text(size = 12),
  ) +
  labs(y = "Density", x = expression(sigma[mu])) +
  ggtitle(expression("Prior Exponential(0.048)"))

prior_exp_1 <- ggplot() +
  xlim(0, 5) +
  ylim(0, 1) +
  geom_function(fun = dexp, args = list(rate = 1), linewidth = 1) +
  theme(
    plot.title = element_text(size = 16),
    axis.text.x = element_text(size = 12, angle = 0),
    axis.text.y = element_text(size = 12, angle = 0),
    axis.title = element_text(size = 12),
  ) +
  labs(y = "Density", x = expression(sigma["w"])) +
  ggtitle(expression("Prior Exponential(1)"))
```

```{r}
plot_grid(prior_normal_50_52, prior_exp_0.048, prior_exp_1)
```

### 4.2 Code the Model on `Stan`

```{stan output.var="spotify_stan_model", language="stan"}
// YOUR CODE HERE
data {                          
  int<lower=1> n;                     //rows in training set
  int<lower=1> num_artist;            //number of artists in training set
  int<lower=1> artist[n];             //artist ID column by song in training set
  vector[n] dance_score;              //danceability scores by song in training set
  vector[n] popularity_score;         //popularity scores by song in training set
}
parameters {
  vector[num_artist] mean_dance;      //vector of posterior danceability score deviations for each artist (size num_artist)
  vector[num_artist] mean_popularity; //vector of posterior popularity score deviations for each artist (size num_artist)
  real overall_mean_dance;            //overall mean in danceability score
  real overall_mean_popularity;       //overall mean in popularity score
  real<lower=0> between_sigma_dance;  //between-artist sd in danceability score
  real<lower=0> between_sigma_popularity; //between-artist sd in popularity score
  real<lower=0> within_sigma_dance;   //within-artist sd in danceability score
  real<lower=0> within_sigma_popularity;  //within-artist sd in popularity score
  real beta;                          //coefficient for danceability score
}
model {
  between_sigma_dance ~ exponential(0.048);     //between-artist sd prior for danceability
  between_sigma_popularity ~ exponential(0.048); //between-artist sd prior for popularity
  within_sigma_dance ~ exponential(1);  //within-artist sd prior for danceability
  within_sigma_popularity ~ exponential(1); //within-artist sd prior for popularity
  overall_mean_dance ~ normal(50, 52);      //overall mean prior for danceability
  overall_mean_popularity ~ normal(50, 52); //overall mean prior for popularity
  beta ~ normal(0, 1);                      //prior for the coefficient of danceability score
  for (j in 1:num_artist){            //danceability and popularity scores priors (deviations per artist)
    mean_dance[j] ~ normal(0, between_sigma_dance);
    mean_popularity[j] ~ normal(0, between_sigma_popularity);
  }
  for (i in 1:n){
    int artist_index = artist[i];     //auxiliary indexing variable
    dance_score[i] ~ normal(overall_mean_dance + mean_dance[artist_index], within_sigma_dance); //likelihood in training set for danceability
    popularity_score[i] ~ normal(overall_mean_popularity + mean_popularity[artist_index] + beta * dance_score[i], within_sigma_popularity); //likelihood in training set for popularity
  }
}

```

### 4.3 Sample and Visualize the Posterior

```{r tags=c()}
levels(spotify_training$artist) <- 1:44

spotify_dictionary <- list(
  n = nrow(spotify_training),
  num_artist = nrow(artist_catalogue),
  artist = as.integer(spotify_training$artist),
  dance_score = spotify_training$danceability,
  popularity_score = spotify_training$popularity
)

# YOUR CODE HERE
posterior_spotify <- sampling(
  object = spotify_stan_model,
  data = spotify_dictionary,
  chains = 4,
  iter = 25000,
  warmup = 5000,
  thin = 20,
  seed = 553,
  cores = getOption("mc.cores", 4)
)

posterior_spotify_sampling <- as.data.frame(posterior_spotify)
```

```{r}
summary_overall <- as.data.frame(summary(posterior_spotify)$summary)
summary_overall <- summary_overall[89:94, c("mean", "sd", "2.5%", "97.5%")] %>%
  mutate_if(is.numeric, round, 3)
summary_overall
```

### Observation

-   The overall `danceability` is between 61.956 and 68.849 with 95% probability and a posterior mean of 65.316.

-   The overall `popularity` is less than danceability between 41.908 and 59.915 with 95% probability and a posterior mean of 51.107. The credible interval is larger because it incorporates the danceability in estimation.

-   Moreover, we see more variability associated to the between-artist standard deviation which align with our assumption. On the other hand, the within-artist variability shows less variability.

```{r}
summary_artist_means <- as.data.frame(summary(posterior_spotify)$summary)
summary_artist_means <- summary_artist_means[45:88, c("mean", "sd", "2.5%", "97.5%")] %>%
  mutate_if(is.numeric, round, 3)
summary_artist_means$artist <- as.factor(artist_catalogue$artist)
summary_artist_means %>%
  arrange(-mean) 
```

```{r}
posterior_artist_means_CIs_plot <- summary_artist_means %>%
  arrange(-mean) %>%
  slice(1:10) %>%
  mutate(artist = fct_reorder(artist, mean)) %>%
  ggplot(aes(x = mean, y = artist)) +
  geom_errorbarh(aes(xmax = `2.5%`, xmin = `97.5%`, color = artist)) +
  geom_point(color = "black") +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.position = "none"
  ) +
  ggtitle("95% CIs for Artist Mean Deviations") +
  labs(x = "Artist Mean Deviation", y = "Artist") +
  geom_vline(xintercept = 0, linetype = "dashed")
```

### See which artist has a higher popularity than the mean

```{r}
posterior_artist_means_CIs_plot
```

From the above plot, we can see the artists that has a higher popularity compared to the overall mean popularity for the songs on the platform, since their credible intervals with positive bounds.

### 4.4 MCMC Diagnosis

```{r fig.width=20, fig.height=30, warning=FALSE, tags=c()}
color_scheme_set("mix-blue-pink")

trace_spotify_4_chains <- mcmc_trace(posterior_spotify,
  pars = c(
    "overall_mean_dance",
    "overall_mean_popularity",
    "between_sigma_dance",
    "between_sigma_popularity",
    "within_sigma_dance",
    "within_sigma_popularity"),
  size = 0.4,
  facet_args = list(nrow = 6)
) +
  ggtitle("Trace Plots by Parameter of Interest") +
  theme(
    plot.title = element_text(size = 16, face = "bold", family = "sans"),
    axis.text = element_text(size = 12, family = "sans"),
    axis.title = element_text(size = 12, family = "sans"),
    legend.text = element_text(size = 12, family = "sans"),
    legend.title = element_text(size = 12, family = "sans")
  ) +
  facet_text(size = 12)
trace_spotify_4_chains 
```

We are seeing a flat pattern by parameter without any upward or a downward trend throughout the chain. Moreover, there isn't any chain stuck. Suggesting the sampling has gone well.

### 4.5 Prediction of all songs (Suppose we do not have any information)

```{stan output.var="spotify_pred_model", language="stan"}
data {                          
  int<lower=1> n;                     
  int<lower=1> num_artist;            
  int<lower=1> artist[n];             
  vector[n] dance_score;              
  vector[n] popularity_score;         
}
parameters {
  vector[num_artist] mean_dance;      
  vector[num_artist] mean_popularity; 
  real overall_mean_dance;            
  real overall_mean_popularity;       
  real<lower=0> between_sigma_dance;  
  real<lower=0> between_sigma_popularity;
  real<lower=0> within_sigma_dance;   
  real<lower=0> within_sigma_popularity;
  real beta;                          
}
model {
  between_sigma_dance ~ exponential(0.048);     
  between_sigma_popularity ~ exponential(0.048);
  within_sigma_dance ~ exponential(1);  
  within_sigma_popularity ~ exponential(1);
  overall_mean_dance ~ normal(50, 52);      
  overall_mean_popularity ~ normal(50, 52);
  beta ~ normal(0, 1);                      
  for (j in 1:num_artist){            
    mean_dance[j] ~ normal(0, between_sigma_dance);
    mean_popularity[j] ~ normal(0, between_sigma_popularity);
  }
  for (i in 1:n){
    int artist_index = artist[i];     
    dance_score[i] ~ normal(overall_mean_dance + mean_dance[artist_index], within_sigma_dance);
    popularity_score[i] ~ normal(overall_mean_popularity + mean_popularity[artist_index] + beta * dance_score[i], within_sigma_popularity);
  }
}
generated quantities {
  vector[n] popularity_score_pred;
  for (i in 1:n){
    int artist_index = artist[i];
    popularity_score_pred[i] = normal_rng(overall_mean_popularity + mean_popularity[artist_index] + beta * dance_score[i], within_sigma_popularity);
  }
}

```

```{r}
posterior_spotify_pred <- sampling(
  object = spotify_pred_model,
  data = spotify_dictionary,
  chains = 1,
  iter = 25000,
  warmup = 5000,
  thin = 20,
  seed = 553
)
```

```{r}
summary_pred <- as.data.frame(summary(posterior_spotify_pred)$summary)
summary_pred  <- summary_pred [96:445,c("mean")]
# Calculate the total sum of squares
sst <- sum((spotify_training$popularity - mean(spotify_training$popularity))^2)

# Calculate the residual sum of squares
ssr <- sum((spotify_training$popularity - summary_pred)^2)

# Calculate the R-squared score
r2_score <- 1 - (ssr / sst)

# Print the R-squared score
print(paste("The R-squared score is", r2_score))
```
